---
title: "tpca_example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tpca_example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(mixOmics)
library(tMixOmics)
library(dplyr)
library(plotly)
```

# Introduction

In this vignette, we will demonstrate how to use the `tpca` function from the `tMixOmics` package. `tpca` is the R implementation of the tensor dimension reduction method described in the paper ["Dimensionality reduction of longitudinal ’omics data using modern tensor factorizations"](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1010212). To demonstrate the method, we will use the data from the first case study of the paper, which is a longitudinal study investigating the reconstitution of the gut microbiome in healthy individuals following antibiotics administration. The data is available in the `tMixOmics` package as `data_Suez`.

# Load the data

The data is already preprocessed and stored in the `data_Suez` data frame. Refer to [Suez2018.ipynb](https://github.com/UriaMorP/tcam_analysis_notebooks/blob/main/Suez2018/Suez2018.ipynb) for details.

```{r}
data("data_Suez")

# convert to factor
data_Suez$rGroup <- factor(data_Suez$rGroup)

data_Suez<-data_Suez%>%
  arrange(rGroup, Participant, rDay)
```

# The tensor data structure

Before we can apply the `tpca`, we must reformat our dataset, that is currently in the form of data frame. While doing so, we must make sure to maintain the information regarding indices for subjects, features (taxa annotations) and timepoints.

```{r}
# Calculate n, p, t
n <- length(unique(data_Suez$Participant)) # number of individuals
p <- ncol(data_Suez)-6 # number of features (exclude first 6 that is the meta data)
t <- length(unique(data_Suez$rDay)) # number of days
#n; p; t

# Initialize clr_array and diet_array
logfold_array <- array(0, dim = c(n, p, t))

# Loop over days
for (i in 1:t) {
  day <- unique(data_Suez$rDay)[i]
  
  # Get index for individuals with the specific day
  (indiv_index <- which(data_Suez$rDay == day))
  
  # Fill array and diet_array
  logfold_array[,,i] <- as.matrix(data_Suez[indiv_index, -c(1:6) ])
}
```

# TCAM results

```{r echo=FALSE}
# Perform TPCA
tpca_res <- tpca(logfold_array,  ncomp = 3)

# Save component scores
  res_df<-data.frame(F1=tpca_res$variates[,1],
                     F2=tpca_res$variates[,2])
  
  #Include Participant ID as rowname and include group details
  rownames(res_df)<-unique(data_Suez$Participant)
  res_df$Group <- data_Suez$rGroup[match(rownames(res_df), data_Suez$Participant)]

# Save loading values and select top 20 loading values
  loadings_cmp1 <- tpca_res$loadings[, 1]
  
  #Sort the loadings based on absolute value
  sorted_indices_cmp1 <- order(abs(loadings_cmp1), decreasing = TRUE)
  sorted_otus_comp1 <- colnames(data_Suez[, -c(1:6) ])[sorted_indices_cmp1[1:20]]
  sorted_loadings_comp1 <- loadings_cmp1[sorted_indices_cmp1[1:20]]

  plot_data <- data.frame(OTUs = sorted_otus_comp1, 
                          Loadings = sorted_loadings_comp1)
  plot_data$OTUs <- factor(plot_data$OTUs, 
                             levels = plot_data$OTUs, ordered = TRUE)
```

## Sample plot

```{r fig1, fig.width=8, fig.height=6, message=FALSE, warning=FALSE}

# Custom color palette
custom_colors <- c("CTR" = "blue", "FMT" = "orange", "PBX" = "green")  # Replace with your actual group names

# Convert res_df into a plotly object with customized colors
plot_ly(data = res_df, 
        x = ~F1, 
        y = ~F2, 
        color = ~Group, 
        colors = custom_colors,  # Use custom colors for groups
        text = rownames(res_df),  # Hover text with row names
        type = 'scatter', 
        mode = 'markers',
        marker = list(size = 15),
        hoverinfo = "text") %>%  # Adjust size of the points
    layout(title = "Tensor PCA",
           xaxis = list(title = "F1"),
           yaxis = list(title = "F2"),
           showlegend = TRUE,
           legend = list(title = list(text = "Group")))

```

## Loading plot 

```{r fig2, fig.width=8, fig.height=6, message=FALSE, warning=FALSE}
plot_ly(data = plot_data,
        x = ~Loadings,
        y = ~OTUs,
        type = 'bar',
        width= 1000,
        height= 400,
        marker = list(color = 'grey'),
  showlegend=F) %>%
  layout(
    title = "Loadings Plot (Component 1)",
    xaxis = list(title = "Loadings"),
    yaxis = list(title = "OTUs"),
    bargap = 0.2  # Adjust the gap between bars if needed
  )
```

